European or American? A re-examination of the transatlantic influence over the IMF

### 2025-09-09 manuscript/manuscript.qmd
spaces before brackets

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

@@

This appendix extends our analysis of IMF conditionality by examining alternative measures that focus specifically on fiscal policy requirements. Following @guimaraes2021, we consider not only the total number of conditions (as in our main analysis), but also the number of *fiscal* conditions and the required fiscal adjustment (RFA)[^14]—the extent of fiscal tightening mandated by IMF programmes, expressed as a percentage of GDP. We report RFA at two time horizons: one year after programme approval ($T+1$) and at the programme’s end year. Together, these measures provide a more nuanced understanding of the nature and stringency of IMF conditionality, particularly in the fiscal domain.

>>>>

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018]. Closer allies of the United States enjoy IMF loans with less fiscal conditions [@dreher2003]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

@@

This appendix extends our analysis of IMF conditionality by examining alternative measures that focus specifically on fiscal policy requirements. Fiscal conditionality, under a comprehensive review of both theoretical approaches and empirical evidence, is "the practice of giving financial assistance contingent on the implementation of specific policies", and it continues to rise over the years [@dreher2009a]. Following @guimaraes2021, we consider not only the total number of conditions (as in our main analysis), but also the number of *fiscal* conditions and the required fiscal adjustment (RFA)[^14]—the extent of fiscal tightening mandated by IMF programmes, expressed as a percentage of GDP. We report RFA at two time horizons: one year after programme approval ($T+1$) and at the programme’s end year. Together, these measures provide a more nuanced understanding of the nature and stringency of IMF conditionality, particularly in the fiscal domain.
